Comparing Early Evolution of Flames in X-ray Bursts in Two and Three Dimensions

### 2023-03-06 xrb3/paper.tex
Comments on first draft of 3D XRB paper

X-ray bursts (XRBs) result from thermonuclear burning of an accreted
H/He or He layer on a neutron star \citep{galloway:2017}.
Observations of brightness oscillations in the rise of the lightcurve
suggest that the burning begins localized and then spreads across the
neutron star \citep{bhattacharyya:2007}.  This spreading is inherently
a multi-dimensional phenomena, and hydrodynamic simulations that
resolve the reactive zone, use realistic reaction networks, and
capture the scales over which rotation is important are needed to
understand the flame propagation and nucleosynthesis (especially
rp-process in mixed H/He bursts, as discussed in \citealt{rpprocess}).
We show a first attempt at modeling the early evolution of a spreading
hotspot in three dimensions.  This builds on our earlier two
dimensional work \citep{eiden:2020,harpole:2021} that developed our
simulation framework and explored how the acceleration of the burning
front depends on the initial model structure.  As with those
simulations, we use the freely-available \castro\ simulation code
\citep{castro,castro_joss}, with all of the code needed to run these
simulations in our public github repos.

Our simulations complement the two- and three-dimensional models of flames of
\citet{cavecchi:2013,art-2015-cavecchi-etal,art-2016-cavecchi-etal,Cavecchi2019}
by focusing on resolving the reaction zone of the flame and modeling
the vertical structure hydrodynamically instead of hydrostatically.
We also focus on modeling flames starting from a small hot spot.  This
however means that we are confined to smaller domains, with similar
computational resources, so one of the goals of this study is to
understand how far we can push resolved XRB flame simulations.  The
work by \citet{goodwin:2021} also explores multidimensional evolution,
looking at the thermal transport and how affects the location where
the hotspot ignites.  Together, all three approaches help build a
comprehensive multi-dimensioanl picture of X-ray bursts.

@@

Moving to 3D is very expensive, so to further reduce the computational
expense, we use an aniostropic resolution, with 32 cm resolution laterally and 16
cm resolution vertically.  This is finer vertically than done in
\citet{harpole:2021}.  Because
the hotspot is placed in the center of the domain, the size of the
domain also becomes a constraint, and as a result, the evolution time
we can reach will become limited by the time it takes the burning to
approach the edge of the domain.  As a result, we focus here on the
early evolution.

@@

\begin{figure}[t]
\centering
%\epsscale{0.75}
\plotone{abar_top_stack}
%\epsscale{1.0}
\caption{\label{fig:vr_abar} Volume rendering of the three-dimensional XRB
simulation showing the mean molecular weight, $\bar{A}$.  The flame is viewed
from above.}
\end{figure}

Figure~\ref{fig:vr_abar} shows the a three-dimensional simulation via
top-down volume rendering of the mean-molecular weight,
\begin{equation}
\frac{1}{\bar{A}} = \sum_k \frac{X_k}{A_k}
\end{equation}
where $X_k$ is the mass fraction of species $k$ and $A_k$ is the
atomic weight (in mass units).  The structure is shown at 10, 20, and
40~ms of simulation time.  While initially quite symmetric, the axial
symmetry is broken and the flame structure takes on a more tenuous
form.  The panels show the flame spreading considerably through the
domain as time evolves, and the simulation is halted once the flame
nears the boundaries.

\begin{figure*}[t]
\centering
\plotone{time_series}
\caption{\label{fig:2d_abar} Time-sequence of the two-dimensional axisymmetric
simualtion using the same setup as the three-dimensional simulation shown
in Figure~\ref{fig:vr_abar}.  The domain is zoomed in, showing only the left
two-thirds of the surface.}
\end{figure*}

\begin{figure*}[t]
\centering
%\epsscale{0.75}
\plotone{flame_wave_1000Hz_25cm_smallplt174562_slice}
%\epsscale{1.0}
\caption{\label{fig:3dslice} A vertical slice through the three-dimensional
simulation showing the entire domain from the side.  The flame structure
is clearly seen.}
\end{figure*}

Figure~\ref{fig:2d_abar} shows the corresponding two-dimensional
axisymmetric simulation.  In the axisymemtric geometry, each zone is
essentially a torus, rotated about the vertical axis.  The initial
model and simulation parameters are identical to the three-dimensional
case.  We see that the overall structure and evolution looks identical
to that reported in \cite{harpole:2021}.  We see that there is a
well-defined flame in the simulation plane that spreads outward with
time.  A vertical slice through the three-dimensional simulation (in
the $x$-$z$ plane) is shown in Figure~\ref{fig:3dslice} at the final
time, 40~ms.  Aside from the lack of reflection symmetry from the
absence of axisymmetric geometry, it looks very similar to the two-dimensional case.

@@

There is still a role for three-dimensional simulations---the
convective burning in the accreted layer should create turbulence that
the flame will encounter as it propagates.  Although we have shown
that for the flame in a Type Ia supernovae, the turbulence from the
era of convection is not strong enough to disrupt the
flame~\citep{wdturb}, the flame in an XRB is considerably slower and
thicker, so it remains to be seen what effect it might have.  This can
be assessed in a different geometry, like a long channel, which would
allow us to trade domain size for resolution to increase the numerical
Reynolds number.  We also will consider higher-order simulation
methodologies which could allow us to capture these effects at lower
resolution.  We've already demonstrated a fully fourth-order accruate
(in space and time) algorithm for coupling hydro, diffusion, and
reactions in \castro~\citep{castro-sdc} that could be used here.  The
main outstanding issue with applying that work to the present problem
is the multilevel time integration.

>>>>

X-ray bursts (XRBs) result from thermonuclear burning of an accreted
H/He or He layer on a neutron star \citep{galloway:2017}.
Observations of brightness oscillations in the rise of the lightcurve
suggest that the burning begins localized and then spreads across the
neutron star \citep{bhattacharyya:2007}. {\color{red} The next sentence
is long and it was hard to read on my initial pass over it. Could consider
splitting the parenthetical note off as a separate sentence, and switching
to active voice (write ``and we need hydrodynamic simulations" and replace
``are needed" with ``in order").} 
This spreading is inherently
a multi-dimensional phenomenon, and hydrodynamic simulations that
resolve the reactive zone, use realistic reaction networks, and
capture the scales over which rotation is important are needed to
understand the flame propagation and nucleosynthesis (especially
rp-process in mixed H/He bursts, as discussed in \citealt{rpprocess}).
{\color{red} Could split paragraph and start next sentence with
"Here we show...".}
We show a first attempt at modeling the early evolution of a spreading
hotspot in three dimensions.  This builds on our earlier two-dimensional
work \citep{eiden:2020,harpole:2021} that developed our
simulation framework and explored how the acceleration of the burning
front depends on the initial model structure.  As with those
simulations, we use the freely-available \castro\ simulation code
\citep{castro,castro_joss}, with all of the code needed to run these
simulations in our public GitHub repositories.

Our simulations complement the two- and three-dimensional models of flames of
\citet{cavecchi:2013,art-2015-cavecchi-etal,art-2016-cavecchi-etal,Cavecchi2019}
by focusing on resolving the reaction zone of the flame and modeling
the vertical structure hydrodynamically instead of hydrostatically.
We also focus on modeling flames starting from a small hot spot.  This
however means that we are confined to smaller domains, with similar
computational resources, so one of the goals of this study is to
understand how far we can push resolved XRB flame simulations.  The
work by \citet{goodwin:2021} also explores multidimensional evolution,
looking at the thermal transport and how affects the location where
the hotspot ignites.  Together, all three approaches help build a
comprehensive multidimensional picture of X-ray bursts.

@@

Moving to 3D is very expensive, so to further reduce the computational
expense, we use an aniostropic resolution, with 32 cm resolution laterally and 16
cm resolution vertically.  This is finer vertically than done in
\citet{harpole:2021}.  Because
the hotspot is placed in the center of the domain, the size of the
domain also becomes a constraint, and as a result, the evolution time
we can reach will become limited by the time it takes the burning to
approach the edge of the domain.  As a result, we focus here on the
early evolution. {\color{red} Very minor: could start previous sentence with
``consequently" or ``therefore" since ``as a result" appears in the previous sentence.}

@@

\begin{figure}[t]
\centering
%\epsscale{0.75}
\plotone{abar_top_stack}
%\epsscale{1.0}
\caption{\label{fig:vr_abar} Volume rendering of the three-dimensional XRB
simulation showing the mean molecular weight, $\bar{A}$.  The flame is viewed
from above. {\color{red} It might be helpful to have the time for each panel either
on the plots or in this caption.}
}
\end{figure}

Figure~\ref{fig:vr_abar} shows the a three-dimensional simulation via
top-down volume rendering of the mean-molecular weight,
\begin{equation}
\frac{1}{\bar{A}} = \sum_k \frac{X_k}{A_k}
\end{equation}
where $X_k$ is the mass fraction of species $k$ and $A_k$ is the
atomic weight (in mass units).  The structure is shown at 10, 20, and
40~ms of simulation time.  While initially quite symmetric, the axial
symmetry is broken and the flame structure takes on a more tenuous
form.  The panels show the flame spreading considerably through the
domain as time evolves, and the simulation is halted once the flame
nears the boundaries.
\begin{figure*}[t]
\centering
\plotone{time_series}
\caption{\label{fig:2d_abar} Time-sequence of the two-dimensional axisymmetric
simulation using the same setup as the three-dimensional simulation shown
in Figure~\ref{fig:vr_abar}.  The domain is zoomed in, showing only the left
two-thirds of the surface.}
\end{figure*}

\begin{figure*}[t]
\centering
%\epsscale{0.75}
\plotone{flame_wave_1000Hz_25cm_smallplt174562_slice}
%\epsscale{1.0}
\caption{\label{fig:3dslice} A vertical slice through the three-dimensional
simulation {\color{red} at 40~ms} showing the entire domain from the side.  The
flame structure is clearly seen. {\color{red} The flame structure could probably
be made a little clearer if this only showed the lower $10^4~\mathrm{cm}$ of the
domain like the 2D plots. Could also try further reducing the peak $\bar{A}$ on the
colorbar (e.g. to 5.5 or 5.25) both here and in the 2D plots to bring out the flame
even more. I think the plots are still pretty clear with the current colorbar though.}}
\end{figure*}

Figure~\ref{fig:2d_abar} shows the corresponding two-dimensional
axisymmetric simulation.  In the axisymmetric geometry, each zone is
essentially a torus, rotated about the vertical axis.  The initial
model and simulation parameters are identical to the three-dimensional
case.  We see that the overall structure and evolution looks identical
to that reported in \cite{harpole:2021}.  We also see that there is a
well-defined flame in the simulation plane that spreads outward with
time.  A vertical slice through the three-dimensional simulation (in
the $x$-$z$ plane) is shown in Figure~\ref{fig:3dslice} at the final
time, 40~ms.  Aside from the lack of reflection symmetry from the
absence of axisymmetric geometry, it looks very similar to the two-dimensional case.

@@

There is still a role for three-dimensional simulations---the
convective burning in the accreted layer should create turbulence that
the flame will encounter as it propagates.  Although we have shown
that for the flame in a Type Ia supernovae, the turbulence from the
era of convection is not strong enough to disrupt the
flame~\citep{wdturb}, the flame in an XRB is considerably slower and
thicker, so it remains to be seen what effect it might have.  This can
be assessed in a different geometry, like a long channel, which would
allow us to trade domain size for resolution to increase the numerical
Reynolds number.  We also will consider higher-order simulation
methodologies which could allow us to capture these effects at lower
resolution.  We've already demonstrated a fully fourth-order accurate
(in space and time) algorithm for coupling hydro, diffusion, and
reactions in \castro~\citep{castro-sdc} that could be used here.  The
main outstanding issue with applying that work to the present problem
is the multilevel time integration.
